Practical Optional Types for Clojure

### 2015-02-28 related-work.tex

\paragraph{Optional type systems}
Reticulated Python~\cite{Vitousek14} is a gradually typed
system for Python, implemented as a source-to-source
translation that inserts dynamic checks at language boundaries
and supporting Python's first-class object system.
Typed Clojure is implemented as a syntactic check on Clojure
code with no option to insert dynamic checks automatically.
Typed Clojure does not support a first-class object system
because Java has nominal classes, however HMaps offer a subset 
of the structural features offered by Reticulated.

>>>>

\paragraph{Other optional and gradual type systems}
In addition to Typed Racket, several other gradual type
systems have been developed recently, targeting existing
dynamically-typed languages.  Reticulated Python~\cite{Vitousek14} is
an experimental gradually typed system for Python, implemented as a
source-to-source translation that inserts dynamic checks at language
boundaries and supporting Python's first-class object system.  Typed
Clojure does not support a first-class object system because Java (and
Clojure) have nominal classes, however HMaps offer an alternative to
the structural objects offered by Reticulated. Similarly,
GradualTalk~\cite{gradualtalk} offers gradual typing for SmallTalk,
with nominal classes.

Optional types, requiring less implementation effort and avoiding any
runtime cost, have been widely adopted in industry, including Hack, an
extension to PHP~\cite{hack}, and Flow~\cite{flow} and
TypeScript~\cite{typescript}, two extensions of JavaScript. These
systems all support some form of occurrence typing, but not in the
generality presented here, nor do they include the other features we
have presented.
